---
title: "R Notebook"
output: html_notebook
---

Cheminformatics abbreviations:
SMILES: simplified molecular-input line-entry system
SDF: Structure Data File


# loading packages

```{r, warning=F, message=F}
library(ChemmineR)
library(rcdk)
library(chemometrics) 
library(cluster)
library(rgl) 
library(tidyverse)
library(vegan) #dissimilarity analysis
library(factoextra) #multivariate data analyses
library(fingerprint) #binary fingerprints of arbitrary length
library(fmcsR) #compute flexible MCS between two sdfs
library(NbClust ) # determining the best number of clusters
library(gridExtra) # Miscellaneous Functions for "Grid" Graphics
library(fpc) # clustering and cluster validation
library(rJava) # simple R-to-Java interface
library(ChemmineOB)  # cheminformatics functionalities OpelBabel C++ project
library(export)
library(scatterplot3d)
library(readxl)
library(data.table)
library(caret)
```


# Convert the smiles format to the sdf format

```{r}
df<-read_excel("Drug compound smiles 1192-21.xls", skip=2) %>% 
  select(1, 8) %>% 
  view()

head(df)

smilist<-df[[2]]

names(smilist)<-df[[1]]

sdfset <- smiles2sdf(smilist)

sdf.visualize(sdfset) 

write.SDF(sdfset, file="DHX36inh_sub.sdf", cid=TRUE)
```

# Visualize thE sdf files

```{r}
# Reading and visualizing sdf files locally 
# https://zinc.docking.org/substances/search/?q=CMP1

sdfset <- read.SDFset(c('DHX36inh_sub.sdf'))

for (i in 1:length(sdfset)) {
  i
  p = plot(CMP[[i]])
  p
}

```
# Visualize the sdfset as a table

```{r}
# data table of compounds including images
# SDFDataTable(sdfset[1:5])
SDFDataTable(sdfset)
htmlwidgets::saveWidget(SDFDataTable(sdfset), "DHX36inh str.html")
```

# Plot a structure of a single SMILES file

```{r}
compound1 <- 'Nc1c(C(Nc(cccc2)c2C(c2ccccc2)=O)=O)sc2ncccc12'
mol1<- parse.smiles(compound1,kekulise=TRUE)[[1]]
mol1
sdf1 <- smiles2sdf(compound1)
plot(sdf1)
```

# Search a compound in the PubChem

```{r}
# Define the smiles of the compound
query <- compound1

# Search
compound_res <- pubchemSmilesSearch(query)

# Plot the top 5 result
for (i in 1:5) {
  i
  p = plot(compound_res[[i]])
  p
}

```


```{r}
# Assigning compound IDs and keeping them unique:
cid(sdfset)[1:50] # Returns IDs from SDFset object
sdfid(sdfset)[1:50] # Returns IDs from SD file header block

unique_ids <- makeUnique(sdfid(sdfset))
cid(sdfset) <- unique_ids 
```
```{r}
# Compute atom frequency matrix, molecular weight and formula:
propma <- data.frame(MF=MF(sdfset), MW=MW(sdfset), atomcountMA(sdfset))
head(propma) 
```

```{r}
# Assign matrix data to data block:
datablock(sdfset) <- propma 
datablock(sdfset[1]) 
```

```{r}
# String searching in the sdfset:
grepSDFset("8011-7983", sdfset, field="datablock", mode="subset") 
grepSDFset("8011-7983", sdfset, field="datablock", mode="index") 
```

```{r}
# Plot molecule structure of one or many SDFs
plot(sdfset[1:4], print=FALSE)
sdf.visualize(sdfset[1:4]) # View compounds in a web browser 
```

```{r}
# Structure similarity searching and clustering:
apset <- sdf2ap(sdfset) ## Generate atom pair descriptor database for searching 
apset

```

```{r}
# Load the apset data 
cmp.search(apset, apset[1], type=3, cutoff = 0.3, quiet=TRUE) ## Search apset database with single compound. 
cmp.cluster(db=apset, cutoff = c(0.65, 0.5), quiet=TRUE)[1:4,] ## Binning clustering using variable similarity cutoffs.
```

```{r}
# OpenBabel Functions
propOB(sdfset[1]) ## Properties from OpenBabel such as 'abonds', 'atoms', 'bonds', 'dbonds', 'HBA1', 'HBA2', 'HBD', 'logP', 'MR', 'MW', 'nF', 'sbonds', 'tbonds', 'TPSA'. 
fingerprintOB(sdfset,"FP2")
```

```{r}
#count rotatable bonds
smartsSearchOB(sdfset[1:5],"[!$(*#*)&!D1]-!@[!$(*#*)&!D1]",uniqueMatches=FALSE)
```

```{r}
# Compute the monoisotopic (exact) mass of a set of compounds
exactMassOB(sdfset[1:5])

sdfset2 = regenerateCoords(sdfset[1:5])

plot(sdfset[4], regenCoords=TRUE,print=FALSE)

openBabelPlot(sdfset[4],regenCoords=TRUE)
```

```{r}
# Generate 3D coordinates for compounds with only 2D coordinates.
sdf3D = generate3DCoords(sdfset[1])
plot(sdf3D)
canonicalSdf = canonicalize(sdfset[1])
mapping = canonicalNumbering(sdfset[1])
```

```{r}
# Define descriptor set in a simple function
desc <- function(sdfset) 
  cbind(SDFID=sdfid(sdfset), 
        MW=MW(sdfset), 
        APFP=desc2fp(x=sdf2ap(sdfset), descnames=1024, type="character"), AP=sdf2ap(sdfset, type="character"), rings(sdfset, type="count", upper=6, arom=TRUE) )  

# Run sdfStream with desc function and write results to a file
sdfStream(input="DHX36inh_sub.sdf", output="matrix.xls", fct=desc, Nlines=1000) 

df_matrix <- fread("matrix.xls")
head(df_matrix)

```

```{r}
# Compound similarity searching
apset <- sdf2ap(sdfset) 
fpset <- desc2fp(apset, descnames=1024, type="FPset") 
fpSim(fpset[1], fpset, method="Tanimoto", cutoff=0.5, top=5) 
```
```{r}
# Clustering according to the distance of fingerprints
simMAap <- sapply(cid(fpset), function(x) fpSim(x=fpset[x], fpset, sorted=FALSE)) 
hc <- hclust(as.dist(1-simMAap), method="single")
plot(as.dendrogram(hc), edgePar=list(col=4, lwd=2), horiz=F)   
```

```{r}
# Clustering according to the atom pair distance matrix
res_cluster <- cmp.cluster(db=apset, cutoff=0, save.distances="distmat.rda", quiet=TRUE) 
load("distmat.rda")
hc <- hclust(as.dist(distmat), method="single") 
hc[["labels"]] <- cid(apset) # Assign correct item labels 
plot(as.dendrogram(hc), edgePar=list(col=4, lwd=2), horiz=F) 
```

```{r}
# heatmap
library(gplots) 
heatmap.2(1-distmat, Rowv=as.dendrogram(hc), Colv=as.dendrogram(hc),col=colorpanel(40, "darkblue", "yellow", "white"), density.info="none", trace="none") 
graph2ppt(file=paste0(Sys.Date(), "_plot",".pptx"), append =TRUE) 
```

```{r}
# Identification of hits from target: mock test
mols <- load.molecules("DHX36inh_sub.sdf") ## limitation: no ID
query.fp <- get.fingerprint(mols[[1]], type = 'maccs')
target.mols <- mols
target.fps <- lapply(target.mols, get.fingerprint, type = 'maccs')
target.fps
sims <-
  data.frame(sim = do.call(
    rbind,
    lapply(
      target.fps,
      fingerprint::distance,
      fp2 = query.fp,
      method = 'tanimoto'
    )
  ))
subset(sims, sim >= 0.7)
hits <- which(sims >= 0.7)
hits
```

```{r}
# visualize the hits
plot(sdfset[hits])
graph2ppt(file=paste0(Sys.Date(), "_plot",".pptx"), append =TRUE) 
```

```{r}
# Distances between CMP1 and the rest of molecules ----
mols<-load.molecules( c('DHX36inh_sub.sdf') ) ## rcdk
query.mol<-mols[[1]]
target.mols<-mols
#target.mols
fps <- lapply(mols, get.fingerprint, type='extended')
fps
fp.sim <- fingerprint::fp.sim.matrix(fps, method='tanimoto')
fp.dist <- 1 - fp.sim
fp.dist
```


```{r}
# Hierarchical clustering ----
d <- dist(fp.dist, method = "euclidean")
hc.res <- hclust(d, method = "ward.D2" )

fviz_dend(hc.res, cex = 0.6, k = 4, color_labels_by_k = TRUE)
graph2ppt(file=paste0(Sys.Date(), "_plot",".pptx"), append =TRUE) 
```

```{r}
# K-means clustering and visualization

# Optimal kmeans cluster number
fviz_nbclust(fp.dist, kmeans, method = "wss") +
geom_vline(xintercept = 4, linetype = 2)

# k-means clustering
km.res <- kmeans(fp.dist, 4, nstart = 5)
km.res
fviz_cluster(km.res, data = fp.dist, ellipse.type = "convex", repel = T)
graph2ppt(file=paste0(Sys.Date(), "_plot",".pptx"), append =TRUE) 

# Statistics for K-means
# stats
km_stats <- cluster.stats(fp.dist,  km.res$cluster)
km_stats

# Silhouette coefficient of observations
sil <- silhouette(km.res$cluster, dist(fp.dist))
head(sil[, 1:3], 10)
plot(sil, main ="Silhouette plot - K-means")
```

```{r}
# Create a 3D scatter plot of MDS result. 
apset <- sdf2ap(sdfset) 
clusters <- cmp.cluster(db=apset, cutoff = c(0.7, 0.8, 0.9), quiet = TRUE)
coord <- cluster.visualize(apset, clusters, size.cutoff=1, dimensions=3, quiet=TRUE) 
scatterplot3d(coord) 
```

```{r}
# Visualize Similarity Search
similarities <- cmp.search(apset, apset[1], type=3, cutoff = 5)
sdf.visualize(sdfset[similarities[,1]]) 
```

# Machine learning

e.g. Predict solubility using an example of dataset below. 

```{r}
# data
data_original <- fread("https://raw.githubusercontent.com/deepchem/deepchem/master/datasets/delaney-processed.csv")
head(data_original)
names(data_original)

library(janitor)
data <- data_original %>% 
  select(-1) %>% 
  janitor::clean_names()

str(data)

# Load necessary libraries
library(randomForest)

# Split data into training and testing sets
set.seed(123)
train_index <- sample(1:nrow(data), 0.7*nrow(data))
train_data <- data[train_index, ]
test_data <- data[-train_index, ]

# Fit random forest model
model <- randomForest(esol_predicted_log_solubility_in_mols_per_litre ~ ., data = train_data)

# Predict bioactivity on test data
predictions <- predict(model, newdata = test_data)

# Evaluate model performance
rmse <- RMSE(predictions, test_data$esol_predicted_log_solubility_in_mols_per_litre)
rmse

# Summary: RMSE, Rsuared, and MAE. 
postResample(pred = predictions, obs = test_data$esol_predicted_log_solubility_in_mols_per_litre)

# plot the prediction
results <- data.frame(Predicted = predictions, Actual = test_data$esol_predicted_log_solubility_in_mols_per_litre)
ggplot(data = results, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Actual", y = "Predicted", title = "Predicted vs Actual")

```

# AutoML method

In this method, a potentially better model could be found by running the  H2O’s AutoML tool and setting up the number of models (max_models). The documents of this tool can be found at https://docs.h2o.ai/h2o/latest-stable/h2o-docs/automl.html. However, the running speed could be slow. 


```{r}
library(h2o)

# Start the H2O cluster (locally)
h2o.init()

# import datasets
train_set2 <- as.h2o(train_data)
test_set2 <- as.h2o(test_data)

# Identify predictors and response
y <- "esol_predicted_log_solubility_in_mols_per_litre"
x <- setdiff(names(train_set2), y)

# Run AutoML for mutliple base models
aml <- h2o.automl(x = x, y = y,
                  training_frame = train_set2,
                  max_models = 5,
                  seed = 1)

# View the AutoML Leaderboard
lb <- aml@leaderboard
print(lb, n = nrow(lb))  # Print all rows instead of default (6 rows)

# The leader model is stored here
fit_aml <- aml@leader
fit_aml

# Retrieve the model performance
perf <- h2o.performance(fit_aml, test_set2)
perf
RMSE_aml <- perf@metrics$RMSE

# Predict solubility on test data
predictions <- predict(fit_aml, newdata = test_set2)$predict %>% 
  as.data.frame() %>% 
  pull()

# plot the prediction
results <- data.frame(Predicted = predictions, Actual = test_data$esol_predicted_log_solubility_in_mols_per_litre)
ggplot(data = results, aes(x = Actual, y = Predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Actual", y = "Predicted", title = "Predicted vs Actual")
```




